biclaR: Estimating the socio-environmental impacts of car substitution by bicycle and public transit using open tools

### 2023-09-27 paper/PaperTRA/PaperTRA.Rmd
finish methods

[^1]: See [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt/)

@@

[^2]: See [pct.bike](https://www.pct.bike/)

@@

[^4]: see [docs.conveyal.com/learn-more/traffic-stress](https://docs.conveyal.com/learn-more/traffic-stress)

@@

## Assessing socio-environmental benefits - CONTINUE FROM HERE

Socio-environmental impacts were estimated for a short term time horizon (i.e., one year) and the long term (i.e., ten years).
using the HEAT for Cycling and the HEAT as a Service[^5] tools, from the WHO.

[^5]: see https://github.com/HEAT-WHO/HEAT_heatr_api

Os impactes foram avaliados para cada cenário e para diferentes escalas territoriais:

-   Desagregado à escala municipal para cada segmento de rede:
    -   Ambientais (emissões de CO~2~eq evitadas);
    -   Balanço monetarizado dos impactes socioeconómicos totais (considerando o balanço dos impactes ambientais e sociais);
-   Agregado à escala metropolitana:
    -   Ambientais (emissões de CO~2~eq evitadas);
    -   Balanço monetarizado dos impactes socioeconómicos totais (considerando o balanço dos impactes ambientais e sociais);
    -   Ambientais (emissões de CO~2~eq, CO, PM10, NOx, e VOC11, evitadas por substituição dos modos motorizados, considerando as emissões geradas na transferência para transportes públicos).

Para todos os cenários, recorreu-se à ferramenta HEAT for cycling, da Organização Mundial de Saúde, para estimativa dos impactes sociais e ambientais da transferência de viagens em automóvel para viagens em bicicleta, nas componentes de: a) Social - Saúde, Atividade Física, Exposição a poluição atmosférica, Exposição ao risco de sinistralidade rodoviária; b) Ambiental - Emissão de gases CO~2~eq.

Additionally, we estimate the impacts of shifting car trips to PT for the second leg of the journey with EMEP/EEA's COPERT methodology and monetize them with the EU Guide to cost-benefit analysis.

A estimativa dos impactes ambientais resulta em toneladas de CO~2~eq, e a estimativa dos impactes sociais resulta em mortalidade prematura evitada.
Ambas as unidades são por fim monetizadas em €, segundo os valores da literatura utilizada em estudos semelhantes.
Para além dos impactes sociais e ambientais resultantes da transferência do automóvel para a bicicleta (na primeira e na última parte da viagem, de e para a interface de TP), estimou-se também o impacte ambiental adicional, resultante da transferência do automóvel para os vários transportes públicos (na segunda etapa da viagem, entre as interfaces).
Como tal, foi necessário caracterizar o universo dos modos motorizados a serem considerados para os cálculos dos respetivos fatores de emissões de gases poluentes e atmosféricos.

Os fatores de consumo e de emissão dos automóveis e transportes públicos deve ser tido em conta relativamente ao número de passageiros transportados (por passageiro.km) e não ao veículo (que seria veículo.km).
No caso do automóvel, para contabilizar a emissão de gases evitados pela transferência de viagens em automóvel para os transportes públicos, teve-se em conta o valor de referência de 1.61 passageiros/automóvel reportados pelo Inquérito à Mobilidade (INE 2017).

As emissões evitadas por cada viagem em transportes públicos que substitui o automóvel correspondem às emissões equivalentes de um automóvel com características correspondentes à média da frota em circulação em Lisboa, para 2 tipos de combustível: gasolina e gasóleo.
Utilizou-se a metodologia e valores de referência do software COPERT (Ntziachristos and Samaras 2020), v5.0, da Agência Europeia do Ambiente, para um nível de detalhe 3 (Tier 3) na estimativa de consumo de emissões para o automóvel, nestes dois tipos de combustível.
Optou-se por usar um veículo de dimensão familiar, norma EURO, e tipo de combustível gasolina ou diesel.
Considerou-se que as viagens foram todas realizadas em condições urbanas (com as respetivas implicações no regime médio de condução) e a uma velocidade média de 15km/h, nos períodos de hora de ponta.
Uma vez que a distância média percorrida por viagem influencia o nível de sobreconsumo e emissões decorrentes da operação do motor a frio -- ou seja, distâncias maiores diluem a importância deste sobreconsumo face ao fator de consumo com o motor a operar a quente, foram estimados os consumos para diferentes gamas de viagem, em intervalos de 500 metros.
As emissões são estimadas para os seguintes poluentes atmosféricos: CO (monóxido de carbono), NO X (Óxidos de Azoto), COV (Compostos Orgânicos Voláteis), PM (material particulado).
Também é estimada as emissões dos principais gases com efeito de estufa: CO2 (dióxido de carbono); CH4 (metano) e N2O (Óxido nitroso), assim como o CO~2~eq equivalente.

Relativamente aos transportes públicos, recorreu-se aos fatores de emissão reportados nos relatórios de sustentabilidade dos respetivos operadores (Carris 2020; Metropolitano de Lisboa 2020; CP 2020, Grupo Transtejo 2014).

a conversão das emissões evitadas em perda de bem-estar evitado, através da respetiva valorização monetária.
a partir dos valores de referência para os vários gases contabilizados, atualizados para 2022 15, com base nas fontes: Bickel et al. (2006), Nash and others (2003), Sartori et al. (2014).

@@

Foram então calculadas as emissões dos poluentes para cada viagem substituída, tendo em conta o modo de transporte, distâncias e velocidade.
A tabela 11 apresenta a estimativa anual de gases emitidos e evitados por transferência do automóvel para os outros modos de transporte público, como exemplo para o cenário com a meta de 4%, usando a rede ciclável "direta".

\[Tabela 11\]

>>>>

[^1]: See [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt/).

@@

[^2]: See [pct.bike](https://www.pct.bike/).

@@

[^4]: see [docs.conveyal.com/learn-more/traffic-stress](https://docs.conveyal.com/learn-more/traffic-stress).

@@

## Assessing socio-environmental benefits

<!-- The impacts were assessed for the total metro area. -->

For the cycling legs of the journey (first and last legs), socio-environmental impacts were estimated, using the HEAT for Cycling tool v5.0 [@HEAT] and the [`HEATaaS` R package](https://github.com/HEAT-WHO/HEAT_heatr_api)[^5] from the World Health Organization.
The HEAT tool provided estimates on the shifting from car to cycling for a short term time horizon (i.e., one year) and the long term (i.e., ten years).
We considered two dimensions: *social* — including the physical activity of cyclists, air pollution exposure, and road casualties; *environmental* — including CO~2~eq emissions and other pollutants.

[^5]: `HEATaaS` is under development. For more information contact [heatwalkingcycling.org](https://heatwalkingcycling.org).

<!-- For the full paper, consider refer the input HEAT values (tabela 6) -->
<!-- Explain why the use of the HaaS tool instead of the website (3 scenarios, 18 municipalities, 2 routing profiles) -->
<!-- Also consider put the FE formula and emmission factor values - see report -->

For the second leg of the journey, we estimate the additional environmental impacts of shifting car trips to PT (between the PT interfaces).

To estimate the car emissions, we used the EMEP/EEA's COPERT software v5 methods and reference values [@COPERT] for a Tier 3 detail level. 
We used a family-size vehicle, EURO standard, and gasoline or diesel fuel.
All trips were considered to be made under urban conditions and at an average speed of 15 km/h during rush hour periods.
Since the average distance traveled per trip influences the overconsumption and emissions from cold-start engine operation, we estimated energy and emission factors for different ranges of trips at 500-meter intervals.
In addition, we assumed an occupancy rate of 1.61 passengers *per* car [@IMOB].

Regarding the PT, we considered the emission factor values reported in the environmental and sustainability reports of the PT operators in the LMA [@Carris2019s; @Metro2019s; @CP2019s; @Transtejo2014].

Emissions were estimated for the following atmospheric pollutants: CO~2~eq, CO, PM10, NOx, and VOC.
In particular, for the urban train and tram -- with 100% electric traction -- only CO~2~eq emissions were considered (resulting from the production of electricity, considering a "well-to-tank" approach), since the other pollutants are not emitted locally.

The conversion of avoided emissions into avoided welfare loss and respective monetary valuation was based on the EU Guide to Cost-benefit Analysis [@EuropeanCommission2014] and the best up-to-date reference values for the various gases [@EuropeanCommission2014; @bickel2006; @UNITE].
The social impacts are in avoided premature mortality. This result is finally monetized using the *Statistical Value of Life* for Portugal [@ANSR2021].
We updated all the monetary reference values of the literature based on the annual inflation rate in Portugal for 2022[^6], and our 10-years estimations assumed a with a discount rate of 5% and inflation of 3%.

[^6]: See [Statistics Portugal tool for inflation rate estimates between years](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ipc).

<!-- For the environmental impacts, using conversion values by the ton, suggested by Bickel et al. (2006), Nash et al. (2003), Sartori et al. (2014) : 8.44 €/ton for CO, 2,867.85 €/ton for NOX, 340,969.27 €/ton for PM10, 7,169.62 €/ton for VOC and 35.85 €/ton for CO2eq. -->

@@

Foram então calculadas as emissões dos poluentes para cada viagem substituída, tendo em conta o modo de transporte, distâncias e velocidade.

### 2023-09-27 paper/PaperTRA/PaperTRA.tex
finish methods

To achieve the cycling targets set by the Portuguese national cycling
strategy for 2025 and 2030 (4\% and 10\%, respectively) {[}2{]}, the
Lisbon's Metropolitan Department of Transport introduced
\emph{biclaR}\footnote{See
  \href{https://biclar.tmlmobilidade.pt/}{biclar.tmlmobilidade.pt}}, a
decision support tool that facilitates the design and development of a
metropolitan cycling network {[}3{]}.

\emph{biclaR} builds on the Propensity to Cycle Tool\footnote{See
  \href{https://www.pct.bike/}{pct.bike}} (PCT), a web application and
research project funded by the UK's Department for Transport in 2015
which launched nationally in 2017 as part of the government's Cycling
and Walking Investment Strategy. The PCT initially used only
origin-destination data for commuting trips as the basis of estimates of
cycling potential at zone, route and route network levels {[}4{]}. The
PCT has been extended to include cycling potential for travel to school
in England {[}5{]} and other trip types in other countries.\footnote{See
  \href{https://www.npt.scot}{npt.scot} and
  \href{https://cruse.bike}{cruse.bike} for examples of the PCT in
  Scotland and Ireland that include estimates of cycling for other
  purposes.} However, to the best of our knowledge, this is the first
time that the method has been integrated with public transport data
using multi-modal routing to estimate the potential and benefits of
multi-stage cycling and PT trips.

@@

The selected route choice algorithm was the
\href{https://ipeagit.github.io/r5r/}{\texttt{r5r} R package} {[}7{]},
which allows for great flexibility in configuring estimated route types,
and which proven to provide most accurate route networks for the city of
Lisbon {[}8{]}. \texttt{r5r} can calculate multi-modal routes using PT
combined with other modes. It enables the identification of the most
direct or safest cycling routes, using the Level of Traffic
Stress\footnote{see
  \href{https://docs.conveyal.com/learn-more/traffic-stress}{docs.conveyal.com/learn-more/traffic-stress}}
(LTS) scale, ranging from 1 to 4, where 1 corresponds to the quietest
(e.g., off-road cycle paths) and 4 corresponds to the least quiet (e.g.,
routes shared with motorized traffic). The routes were estimated for the
base scenario for both types of networks: \emph{direct} and \emph{safe},
using LTS 4 and LTS 3, respectively.

@@

\hypertarget{assessing-socio-environmental-benefits---continue-from-here}{%
\subsection{Assessing socio-environmental benefits - CONTINUE FROM
HERE}\label{assessing-socio-environmental-benefits---continue-from-here}}

Socio-environmental impacts were estimated for a short term time horizon
(i.e., one year) and the long term (i.e., ten years). using the HEAT for
Cycling and the HEAT as a Service\footnote{see
  \url{https://github.com/HEAT-WHO/HEAT_heatr_api}} tools, from the WHO.

Os impactes foram avaliados para cada cenário e para diferentes escalas
territoriais:

\begin{itemize}
\tightlist
\item
  Desagregado à escala municipal para cada segmento de rede:

  \begin{itemize}
  \tightlist
  \item
    Ambientais (emissões de CO\textsubscript{2}eq evitadas);
  \item
    Balanço monetarizado dos impactes socioeconómicos totais
    (considerando o balanço dos impactes ambientais e sociais);
  \end{itemize}
\item
  Agregado à escala metropolitana:

  \begin{itemize}
  \tightlist
  \item
    Ambientais (emissões de CO\textsubscript{2}eq evitadas);
  \item
    Balanço monetarizado dos impactes socioeconómicos totais
    (considerando o balanço dos impactes ambientais e sociais);
  \item
    Ambientais (emissões de CO\textsubscript{2}eq, CO, PM10, NOx, e
    VOC11, evitadas por substituição dos modos motorizados, considerando
    as emissões geradas na transferência para transportes públicos).
  \end{itemize}
\end{itemize}

Para todos os cenários, recorreu-se à ferramenta HEAT for cycling, da
Organização Mundial de Saúde, para estimativa dos impactes sociais e
ambientais da transferência de viagens em automóvel para viagens em
bicicleta, nas componentes de: a) Social - Saúde, Atividade Física,
Exposição a poluição atmosférica, Exposição ao risco de sinistralidade
rodoviária; b) Ambiental - Emissão de gases CO\textsubscript{2}eq.

Additionally, we estimate the impacts of shifting car trips to PT for
the second leg of the journey with EMEP/EEA's COPERT methodology and
monetize them with the EU Guide to cost-benefit analysis.

A estimativa dos impactes ambientais resulta em toneladas de
CO\textsubscript{2}eq, e a estimativa dos impactes sociais resulta em
mortalidade prematura evitada. Ambas as unidades são por fim monetizadas
em €, segundo os valores da literatura utilizada em estudos semelhantes.
Para além dos impactes sociais e ambientais resultantes da transferência
do automóvel para a bicicleta (na primeira e na última parte da viagem,
de e para a interface de TP), estimou-se também o impacte ambiental
adicional, resultante da transferência do automóvel para os vários
transportes públicos (na segunda etapa da viagem, entre as interfaces).
Como tal, foi necessário caracterizar o universo dos modos motorizados a
serem considerados para os cálculos dos respetivos fatores de emissões
de gases poluentes e atmosféricos.

Os fatores de consumo e de emissão dos automóveis e transportes públicos
deve ser tido em conta relativamente ao número de passageiros
transportados (por passageiro.km) e não ao veículo (que seria
veículo.km). No caso do automóvel, para contabilizar a emissão de gases
evitados pela transferência de viagens em automóvel para os transportes
públicos, teve-se em conta o valor de referência de 1.61
passageiros/automóvel reportados pelo Inquérito à Mobilidade (INE 2017).

As emissões evitadas por cada viagem em transportes públicos que
substitui o automóvel correspondem às emissões equivalentes de um
automóvel com características correspondentes à média da frota em
circulação em Lisboa, para 2 tipos de combustível: gasolina e gasóleo.
Utilizou-se a metodologia e valores de referência do software COPERT
(Ntziachristos and Samaras 2020), v5.0, da Agência Europeia do Ambiente,
para um nível de detalhe 3 (Tier 3) na estimativa de consumo de emissões
para o automóvel, nestes dois tipos de combustível. Optou-se por usar um
veículo de dimensão familiar, norma EURO, e tipo de combustível gasolina
ou diesel. Considerou-se que as viagens foram todas realizadas em
condições urbanas (com as respetivas implicações no regime médio de
condução) e a uma velocidade média de 15km/h, nos períodos de hora de
ponta. Uma vez que a distância média percorrida por viagem influencia o
nível de sobreconsumo e emissões decorrentes da operação do motor a frio
-- ou seja, distâncias maiores diluem a importância deste sobreconsumo
face ao fator de consumo com o motor a operar a quente, foram estimados
os consumos para diferentes gamas de viagem, em intervalos de 500
metros. As emissões são estimadas para os seguintes poluentes
atmosféricos: CO (monóxido de carbono), NO X (Óxidos de Azoto), COV
(Compostos Orgânicos Voláteis), PM (material particulado). Também é
estimada as emissões dos principais gases com efeito de estufa: CO2
(dióxido de carbono); CH4 (metano) e N2O (Óxido nitroso), assim como o
CO\textsubscript{2}eq equivalente.

Relativamente aos transportes públicos, recorreu-se aos fatores de
emissão reportados nos relatórios de sustentabilidade dos respetivos
operadores (Carris 2020; Metropolitano de Lisboa 2020; CP 2020, Grupo
Transtejo 2014).

a conversão das emissões evitadas em perda de bem-estar evitado, através
da respetiva valorização monetária. a partir dos valores de referência
para os vários gases contabilizados, atualizados para 2022 15, com base
nas fontes: Bickel et al.~(2006), Nash and others (2003), Sartori et
al.~(2014).

@@

Foram então calculadas as emissões dos poluentes para cada viagem
substituída, tendo em conta o modo de transporte, distâncias e
velocidade. A tabela 11 apresenta a estimativa anual de gases emitidos e
evitados por transferência do automóvel para os outros modos de
transporte público, como exemplo para o cenário com a meta de 4\%,
usando a rede ciclável ``direta''.

\[Tabela 11\]

@@

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-IMOB}{}}%
\CSLLeftMargin{1. }%
\CSLRightInline{INE: Mobilidade e funcionalidade do território nas
{Áreas Metropolitanas do Porto e de Lisboa}: 2017,
\url{https://www.ine.pt/xportal/xmain?xpid=INE\&xpgid=ine_publicacoes\&PUBLICACOESpub_boui=349495406\&PUBLICACOESmodo=2\&xlang=pt},
(2018).}

@@

>>>>

To achieve the cycling targets set by the Portuguese national cycling
strategy for 2025 and 2030 (4\% and 10\%, respectively) {[}2{]}, the
Lisbon's Metropolitan Department of Transport introduced
\emph{biclaR}\footnote{See
  \href{https://biclar.tmlmobilidade.pt/}{biclar.tmlmobilidade.pt}.}, a
decision support tool that facilitates the design and development of a
metropolitan cycling network {[}3{]}.

\emph{biclaR} builds on the Propensity to Cycle Tool\footnote{See
  \href{https://www.pct.bike/}{pct.bike}.} (PCT), a web application and
research project funded by the UK's Department for Transport in 2015
which launched nationally in 2017 as part of the government's Cycling
and Walking Investment Strategy. The PCT initially used only
origin-destination data for commuting trips as the basis of estimates of
cycling potential at zone, route and route network levels {[}4{]}. The
PCT has been extended to include cycling potential for travel to school
in England {[}5{]} and other trip types in other countries.\footnote{See
  \href{https://www.npt.scot}{npt.scot} and
  \href{https://cruse.bike}{cruse.bike} for examples of the PCT in
  Scotland and Ireland that include estimates of cycling for other
  purposes.} However, to the best of our knowledge, this is the first
time that the method has been integrated with public transport data
using multi-modal routing to estimate the potential and benefits of
multi-stage cycling and PT trips.

@@

The selected route choice algorithm was the
\href{https://ipeagit.github.io/r5r/}{\texttt{r5r} R package} {[}7{]},
which allows for great flexibility in configuring estimated route types,
and which proven to provide most accurate route networks for the city of
Lisbon {[}8{]}. \texttt{r5r} can calculate multi-modal routes using PT
combined with other modes. It enables the identification of the most
direct or safest cycling routes, using the Level of Traffic
Stress\footnote{see
  \href{https://docs.conveyal.com/learn-more/traffic-stress}{docs.conveyal.com/learn-more/traffic-stress}.}
(LTS) scale, ranging from 1 to 4, where 1 corresponds to the quietest
(e.g., off-road cycle paths) and 4 corresponds to the least quiet (e.g.,
routes shared with motorized traffic). The routes were estimated for the
base scenario for both types of networks: \emph{direct} and \emph{safe},
using LTS 4 and LTS 3, respectively.

@@

\hypertarget{assessing-socio-environmental-benefits}{%
\subsection{Assessing socio-environmental
benefits}\label{assessing-socio-environmental-benefits}}

For the cycling legs of the journey (first and last legs),
socio-environmental impacts were estimated, using the HEAT for Cycling
tool v5.0 {[}9{]} and the
\href{https://github.com/HEAT-WHO/HEAT_heatr_api}{\texttt{HEATaaS} R
package}\footnote{\texttt{HEATaaS} is under development. For more
  information contact
  \href{https://heatwalkingcycling.org}{heatwalkingcycling.org}.} from
the World Health Organization. The HEAT tool provided estimates on the
shifting from car to cycling for a short term time horizon (i.e., one
year) and the long term (i.e., ten years). We considered two dimensions:
\emph{social} --- including the physical activity of cyclists, air
pollution exposure, and road casualties; \emph{environmental} ---
including CO\textsubscript{2}eq emissions and other pollutants.

For the second leg of the journey, we estimate the additional
environmental impacts of shifting car trips to PT (between the PT
interfaces).

To estimate the car emissions, we used the EMEP/EEA's COPERT software v5
methods and reference values {[}10{]} for a Tier 3 detail level. We used
a family-size vehicle, EURO standard, and gasoline or diesel fuel. All
trips were considered to be made under urban conditions and at an
average speed of 15 km/h during rush hour periods. Since the average
distance traveled per trip influences the overconsumption and emissions
from cold-start engine operation, we estimated energy and emission
factors for different ranges of trips at 500-meter intervals. In
addition, we assumed an occupancy rate of 1.61 passengers \emph{per} car
{[}1{]}.

Regarding the PT, we considered the emission factor values reported in
the environmental and sustainability reports of the PT operators in the
LMA {[}11--14{]}.

Emissions were estimated for the following atmospheric pollutants:
CO\textsubscript{2}eq, CO, PM10, NOx, and VOC. In particular, for the
urban train and tram -- with 100\% electric traction -- only
CO\textsubscript{2}eq emissions were considered (resulting from the
production of electricity, considering a ``well-to-tank'' approach),
since the other pollutants are not emitted locally.

The conversion of avoided emissions into avoided welfare loss and
respective monetary valuation was based on the EU Guide to Cost-benefit
Analysis {[}15{]} and the best up-to-date reference values for the
various gases {[}15--17{]}. The social impacts are in avoided premature
mortality. This result is finally monetized using the \emph{Statistical
Value of Life} for Portugal {[}18{]}. We updated all the monetary
reference values of the literature based on the annual inflation rate in
Portugal for 2022\footnote{See
  \href{https://www.ine.pt/xportal/xmain?xpid=INE\&xpgid=ipc}{Statistics
  Portugal tool for inflation rate estimates between years}.}, and our
10-years estimations assumed a with a discount rate of 5\% and inflation
of 3\%.

@@

Foram então calculadas as emissões dos poluentes para cada viagem
substituída, tendo em conta o modo de transporte, distâncias e
velocidade.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-IMOB}{}}%
\CSLLeftMargin{1. }%
\CSLRightInline{INE:
\href{https://www.ine.pt/xportal/xmain?xpid=INE\&xpgid=ine_publicacoes\&PUBLICACOESpub_boui=349495406\&PUBLICACOESmodo=2\&xlang=pt}{Mobilidade
e funcionalidade do território nas {Áreas Metropolitanas do Porto e de
Lisboa}: 2017}. {Instituto National de Estatística}, Lisboa (2018).}

@@

\leavevmode\vadjust pre{\hypertarget{ref-HEAT}{}}%
\CSLLeftMargin{9. }%
\CSLRightInline{Kahlmeier, S., Götschi, T., Cavill, N., Castro
Fernandez, A., Brand, C., Rojas Rueda, D., Woodcock, J., Kelly, P.,
Lieb, C., Oja, P., others:
\href{https://www.euro.who.int/__data/assets/pdf_file/0010/352963/Heat.pdf}{Health
economic assessment tool ({HEAT}) for walking and for cycling: Methods
and user guide on physical activity, air pollution, injuries and carbon
impact assessments}. (2017).}

\leavevmode\vadjust pre{\hypertarget{ref-COPERT}{}}%
\CSLLeftMargin{10. }%
\CSLRightInline{Ntziachristos, L., Samaras, Z.: {EMEP/EEA} air pollutant
emission inventory guidebook 2019,
\url{https://www.emisia.com/utilities/copert/documentation/}, (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-Carris2019s}{}}%
\CSLLeftMargin{11. }%
\CSLRightInline{Carris:
\href{https://www.carris.pt/media/dkdp2wbg/dnf_carris2019_rv5.pdf}{Relatório
de sustentabilidade 2019 - demostração não financeira}. {Carris -
Companhia Carris de Ferro de Lisboa, E.M., S.A.} (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-Metro2019s}{}}%
\CSLLeftMargin{12. }%
\CSLRightInline{Metropolitano de Lisboa:
\href{https://www.metrolisboa.pt/wp-content/uploads/2021/01/relatorio_integrado_2019.pdf}{Relatório
integrado 2019}. {Metropolitano de Lisboa, E.P.E.} (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-CP2019s}{}}%
\CSLLeftMargin{13. }%
\CSLRightInline{CP:
\href{https://www.cp.pt/StaticFiles/Institucional/2_gestao_sustentavel/1_RelatoriosSustentabilidade/relatorio-de-sustentabilidade-2019.pdf}{Relatório
de sustentabilidade 2019}. {CP - Comboios de Portugal, E.P.E.} (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-Transtejo2014}{}}%
\CSLLeftMargin{14. }%
\CSLRightInline{Transtejo:
\href{https://ttsl.pt/wp-content/uploads/2018/01/rs_2014_min.pdf}{Relatório
de sustentabilidade 2014}. {Grupo Transtejo, S.A.} (2014).}

\leavevmode\vadjust pre{\hypertarget{ref-EuropeanCommission2014}{}}%
\CSLLeftMargin{15. }%
\CSLRightInline{Sartori, D., Catalano, G., Genco, M., Pancotti, C.,
Sirtori, E., Vignetti, S., Del Bo, C.:
\href{https://ec.europa.eu/regional_policy/sources/docgener/studies/pdf/cba_guide.pdf}{Guide
to cost-benefit analysis of investment projects. Economic apraisal tool
for cohesion policy 2014-2020}. {European Commission - Directorate
General for Regional and Urban Policy} (2014).}

\leavevmode\vadjust pre{\hypertarget{ref-bickel2006}{}}%
\CSLLeftMargin{16. }%
\CSLRightInline{Bickel, P., Friedrich, R., Burgess, A., Fagiani, P.,
Hunt, A., Jong, G.D., Laird, J.:
\href{https://trimis.ec.europa.eu/sites/default/files/project/documents/20130122_113653_88902_HEATCO_D5_summary.pdf}{{HEATCO
- Developing Harmonised European Approaches for Transport Costing and
Project Assessment. Deliverable 5, Proposal for Harmonised Guidelines.}}
(2006).}

\leavevmode\vadjust pre{\hypertarget{ref-UNITE}{}}%
\CSLLeftMargin{17. }%
\CSLRightInline{Nash, C., others:
\href{https://www.its.leeds.ac.uk/projects/unite/downloads/Unite\%20Final\%20Report.pdf}{UNIfication
of accounts and marginal costs for transport efficiency: Final report}.
{Institute for Transport Studies, University of Leeds} (2003).}

\leavevmode\vadjust pre{\hypertarget{ref-ANSR2021}{}}%
\CSLLeftMargin{18. }%
\CSLRightInline{Silva, C., Bravo, J.V., Gonçalves, J.:
\href{http://www.ansr.pt/Estatisticas/RelatoriosTematicos/Documents/O\%20Impacto\%20Economico\%20e\%20Social\%20da\%20Sinistralidade\%20-\%20PT.pdf}{{Impacto
Económico e Social da Sinistralidade Rodoviária em Portugal}}. {CEGE -
Centro de Estudos de Gestão do ISEG, Autoridade Nacional de Segurança
Rodoviária (ANSR)} (2021).}
